Efficient implementation of penalized regression\\for genetic risk prediction

### 2018-10-11 paper/PRS-supp.tex

We used three different ways to estimate the maximum achievable AUC for our simulations.

First, we used the estimation from equation (3) of \cite{wray2010genetic}. For a prevalence fixed at 30\% and an heritability of 50\% (respectively 80\%), the approximated theoretical values of AUC are 84.1\% (respectively 93.0\%). Note that this approximation is reported to be less accurate for high heritabilities.

Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ through Monte Carlo simulations. We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.

Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table 2), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates. For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
\caption{Comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%. Mean of AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max''). Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-logit}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%. Mean of AUC over 100 simulations. Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-PRS}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
\caption{Comparison of models in scenario \textnumero1 for model ``ADD''. Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-all-r2}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''. Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}). Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-chr6-all-r2}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
\caption{Comparison of T-Trees and PLR in scenario \textnumero1 for an heritability of 80\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean of AUC over 5 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 5 simulations. \textbf{C:} Boxplots of execution times for 5 simulations.}
\label{fig:supp-ttrees}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-triple}}
\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 80\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean of AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations. \textbf{C:} Boxplots of execution times for 100 simulations.}
\label{fig:supp-triple}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 50\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean of AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-triple}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
\caption{Comparison of PLR and the best prediction (among 100 tested $\lambda$ values) for ``biglasso'' (another implementation of penalized logistic regression -- \cite{zeng2017efficient}) in scenario \textnumero1. Simulations use model ``ADD'', an heritability of 80\% and $\alpha = 1$. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. \textbf{A:} Mean of AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations. \textbf{C:} Boxplots of execution times for 100 simulations.}
\label{fig:supp-biglasso}
\end{figure}

>>>>

We use three different ways to estimate the maximum achievable AUC for simulations (see supplementary notebook ``oracle'').

First, we use the estimation from equation (3) of \cite{wray2010genetic}. For a prevalence fixed at 30\% and an heritability of 50\% (respectively 80\%), the approximated theoretical values of AUC are 84.1\% (respectively 93.0\%). Note that this approximation is reported to be less accurate for high heritabilities.

Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ (and prevalence $K$) through Monte Carlo simulations. We report AUCs of 84.1\% and 94.1\% for heritabilities of 50\% and 80\%, respectively.

Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table 2), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities, through 100 replicates. For every combination of parameters, AUC values of oracles vary between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
\caption{Comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%. Mean AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max''). Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-logit}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%. Mean AUC over 100 simulations. Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-PRS}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
\caption{Comparison of methods in scenario \textnumero1 for model ``ADD''. Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-all-r2}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''. Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}). Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-chr6-all-r2}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
\caption{Comparison of T-Trees and PLR in scenario \textnumero1 for an heritability of 80\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean AUC over 5 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 5 simulations. \textbf{C:} Boxplots of execution times for 5 simulations.}
\label{fig:supp-ttrees}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-triple}}
\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 80\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations. \textbf{C:} Boxplots of execution times for 100 simulations.}
\label{fig:supp-triple}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 50\%. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes. \textbf{A:} Mean AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-triple}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
\caption{Comparison of PLR and the best prediction (among 100 tested $\lambda$ values) for ``biglasso'' (another implementation of penalized logistic regression -- \cite{zeng2017efficient}) in scenario \textnumero1. Simulations use model ``ADD'', an heritability of 80\% and $\alpha = 1$. Vertical panels are presenting results for effects following a Gaussian or Laplace distribution. \textbf{A:} Mean AUC over 100 simulations. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC. \textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations. \textbf{C:} Boxplots of execution times for 100 simulations.}
\label{fig:supp-biglasso}
\end{figure}
